---
title: "Scoring records by topic relevance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring records by topic relevance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Chunks run a Python model or read a cache, so they are shown but not
  # executed when the vignette is built.
  eval = FALSE
)
library(planscanR.screen)
```

`planscanR.screen` scores a tibble of text records by how relevant they are to
topics you care about. It is the general-purpose screening framework of the
planscanR family: records usually come from `planscanR::get_assessments()` or
`planscanR::index_cache()`, but any tibble with `title` / `summary` columns
works.

## Score by relevance

Pass a `topic` and each record's title and summary are compared against it with
a multilingual embedding model, producing a relevance score between -1 and 1
(higher means a closer match). A topic is just a short phrase; pass one, or
several named ones:

```{r}
records <- planscanR::index_cache(country = "nl")
records <- score_assessments(
  records,
  topic = c(wind = "wind energy", solar = "solar energy")
)
```

This adds `relevance_score_wind` and `relevance_score_solar` columns. Because
the comparison uses a multilingual model, an English topic phrase still matches
Dutch or German text — you do not need to translate your topics per country.

The embedding model is pluggable. The built-in backend is
`embedding_model_minilm()` (sentence-transformers MiniLM, via reticulate); wrap
any embedding function with `embedding_model()` to use a different one.

> **One-time setup.** The MiniLM backend runs a small Python model through
> reticulate; the package declares `sentence-transformers` via
> `reticulate::py_require()` on load, so it is provisioned on first use.

## Classify and keyword-score

Two complementary signals layer on top of the cosine score. Both take their
vocabulary as an explicit argument — the framework ships no project defaults, so
you define the candidate labels and keyword terms yourself:

```{r}
# Zero-shot classification against a labelled set of candidate classes. Names
# are stable slugs (used as column suffixes); values are the hypotheses fed to
# the model. Include explicit negative classes to catch look-alikes.
labels <- c(
  wind  = "wind energy project",
  solar = "solar energy project",
  other = "not about renewable energy"
)
records <- classify_assessments(records, labels = labels)

# A transparent lexical (substring) count over a multilingual lexicon: a named
# list of term vectors, one per topic.
lexicon <- list(
  wind  = c("wind", "turbine", "windpark", "windmolen"),
  solar = c("solar", "zonne", "photovolta", "pv")
)
records <- score_keywords(records, lexicon = lexicon)
```

## Learn a selection model

Given human keep/drop labels (e.g. exported from a review tool), you can train a
selection model over the per-record scores instead of hand-tuning a rule. Pass
the same `topics` and `labels` you scored and classified with, so the feature
columns line up:

```{r}
model <- train_selection_model(
  records,
  reviews,
  topics = c(wind = "wind energy", solar = "solar energy"),
  labels = labels
)
records <- predict_selection(model, records)
```

The model stores its `topics`/`labels` so `predict_selection()` rebuilds exactly
the same feature frame it was trained on (no train/serve skew).

## Where to go next

* `?score_assessments`, `?score_records` — relevance scoring options.
* `?classify_assessments`, `?classify_model_zeroshot` — the classifier.
* `?train_selection_model`, `?selection_features` — the learned selection model.
* `?embedding_model`, `?classifier`, `?selection_learner` — plug in your own
  backends.
